Modal Abstractions for Virtualizing Memory Addresses

### 2025-03-25 examples.tex

If the entry is not set, \textsf{pte\_initialize} (Line \ref{line:call_to_pte_initialize} in Figure \ref{fig:calltopteinitialize}) 
allocates a physical page (internally utilizing the only unverified (trusted) code in our case studies, the page-allocator's \textsf{kalloc},\footnote{
  This is an allocator for regions of pre-zeroed physical memory that is mapped, but not accessed by the allocator itself,
  as is typical for slab allocators~\cite{bonwick1994slab}.
  Its verification would be similar to verifying a usermode \textsf{malloc} verifications~\cite{Chlipala2013Bedrock,wickerson2010explicit},
  just with additional invariants on the memory pool.
} 
on Line \ref{line:call_to_kalloc} in Figure \ref{pteinitializespec}). 
\looseness=-1
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{\_} \ast \lnot(\textsf{entry\_present entry\_val} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
;;void pte_initialize(pte_t *entry) { ... Stack frame setup elided    
callq  81 <kalloc> ;;allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kalloc}*)
mov    rax,-0x10[rbp] ;; Store into 'local'
;;entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);
mov    -0x10[rbp],rax; mov    -0x8[rbp],rdi; mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \lnot(\textsf{entry\_present entry\_val}) }$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
$\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$
... ;;clean up the stack, return
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\vspace{-1em}
\end{figure}

Since we are using \textsf{pte\_initialize} for page-table address allocation, we must relate this newly
allocated physical address to the identity mapping map $\Xi$ --- 
see Line \ref{line:page_of_caps} in Figure \ref{fig:calltopteinitialize}, where
\texttt{kalloc}'s specification guarantees it has returned memory from a designated memory
pool that is already mapped
\ifPLDI
\else
\footnote{A reasonable reader might wonder where this pool
initially comes from, and how it might grow when needed. Typically an initial mapping subject to this identity mapping
constraint is set up prior to transition to 64-bit kernel code (notably,
a page table must exist \emph{before} virtual memory is enabled during boot, as part of enabling it is setting
a page table root).
Growing this pool later requires cooperation of physical memory range allocation and virtual memory range allocation,
typically by starting general virtual address allocation at the highest physical memory address plus the identity mapping offset.
This reserves the virtual addresses corresponding to all physical addresses plus the offset for later use in this pool,
as needed.
} 
\fi
and satisfies the offset invariants.
% \todo[inline,color=blue]{colin frontier.
% Stuck with line 31 onwards in Figure 7. rax holds nextpaddr, but I think that should be entrypfn, and 
% the explicit entrypfn id token assertion should go away, as its covered by the forall assertion.
% then the postcondition for pte-initialize should have a specific level now for the entries,
% like 0, which can be updated in the view shift on line 42.
% }
% Focusing on the specification of \textsf{pte\_initialize} separately in Figure \ref{fig:pteinitializespec}, 
% we right immediately realize that instead of seeing see a physical pointsto for the fresly page-table address 
% (e.g. $\mathsf{nextpaddr} \mapsto_{\mathsf{p}} \mathsf{w64\_0}$) deliberately in the post-conditoin in Lines 15-16,
%  we observe a full-ownership token representing the knowledge that a frame and all the entries indexed from this 
% frame are freshly allocated with full-ownership to be a part of the identity map, $\Xi$. 
The soundness argument of this specification relies on the fact that these freshly allocated resources are part 
of an entry construction that has not been completed yet: the presence bit is set 
(Line \ref{line:install_new_entry} in Figure \ref{fig:calltopteinitialize}) after these freshly allocated resources are incorporated to the 
entry construction via the page-frame portion of the PTE. In other words, the side condition, (*),
 formalizes that any access to the entry with these resources is \textit{invalid} (in the sense of not necessarily
having accompanying resources) until the entry is marked present (and thus the memory returned from \textsf{kalloc}
moves into the page table invariant.

@@

Realizing a software page-table walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdir}. 
The special part of the specficication for a page table walk can be considered as accumulation of memory mappings for the page-table entries 
visited and frame addresses for page-tables. 
For example, Lines \ref{line:ex_l4_vpte} and \ref{line:ex_l3_vpte} in Figure \ref{walkpgdir} show the virtual pte-pointsto assertions for L4 and L3 entries.
In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk}}$ -- 
which allows us to construct and return the path to the L1 entry in the tree to insert a new page.

@@

\subsection{Mapping a New Page}
\label{sec:mapnew}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table via a call
\[\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t fpaddr, uint64\_t perm )}\] in Figure \ref{fig:mapping_code}.
To do so, with a given allocated a fresh page (\textsf{fpaddr}), then calculate the appropriate
known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdir} in Figure \ref{fig:mapping_code})  and update 
the appropriate L1 page table entry (Line 11 in Figure \ref{fig:mapping_code});
unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t fpaddr, uint64_t perm ) {
... ;;setting up the stack      
 $\specline{\textsf{P} \ast \textsf{pml4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$      
;; pte_t *pte = walkpgdir(pml4, a);
mov    -0x8[rbp],rdi; mov    -0x28[rbp],rsi; mov    0x0,rax; callq  291 <walkpgdir> (*\label{line:call_walkpgdir}*)
;;if(pte->present != 0) {
mov    0x0,r8; mov    rax,rcx; mov    rcx,-0x30[rbp]
... ;; Copy user access bit, writethrough, etc. (not part of our semantics) into pte from perm
\ghostmaptoken{\textsf{id}}{(\mathsf{pml4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm2\_entry})}{\textsf{2}}  \\ \forall_{i\in\textsf{0..511}} \; \textsf{table\_root(pml4\_entry}_{\textsf{pfn}}) \textsf{ + i * 8} \mapsto_{\textsf{id}} \textsf{3}  \\ \forall_{i\in\textsf{0..511}} \; \textsf{table\_root(pdp\_entry}_{\textsf{pfn}}) \textsf{ + i * 8} \mapsto_{\textsf{id}} \textsf{2}  \\ \forall_{i\in\textsf{0..511}} \; \textsf{table\_root(pd\_entry}_{\textsf{pfn}}) \textsf{ + i * 8} \mapsto_{\textsf{id}} \textsf{1}  \\  \textsf{pml4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4e\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l4e\_val})\urcorner \\  \textsf{pdp\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3e\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l3e\_val})\urcorner \\  \textsf{pd\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2e\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l2e\_val})\urcorner \\   \textsf{pt\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l1e\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l1e\_val})\urcorner    \end{array} \right)  =\textsf{R}_{\textsf{walk\_acc}}}$
;;pte->pfn          = PTE_ADDR_TO_PFN(fpaddr);
mov    -0x18[rbp],rcx; shr    0xc,rcx; mov    -0x30[rbp],rdx; mov    [rdx],rsi ;; Load current entry in rsi
$\specline{\textsf{pt\_entry}_{\textsf{pfn}} \mapsto_{p} \textsf{addr\_to\_pfn}(\textsf{pa}) \ast \textsf{R}_{\textsf{walk\_acc}}}_{\rtv} \Rrightarrow \specline{\textsf{va} \mapsto_{\textsf{v,q}} \textsf{pa} \ast \textsf{P}}$
movabs 0xffffffffff,rdi; and    rdi,rcx; shl    0xc,rcx; movabs 0xfff0000000000fff,rdi
and    rdi,rsi; or     rcx,rsi; mov    rsi,[rdx] ;; Store updated entry back to L1 entry (*\label{line:l1entry_store}*)
;;}
... ;; set present bit in entry, then clean up the stack and return (proof re-packs invariants)
;;}
  \caption{Specification and proof of code for mapping a new page with $\textsf{R}_{\textsf{walk\_acc}}$ in Figure \ref{walkpgdir} expanded.}
\label{fig:mapping_code}

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline{walkpgdir} (discussed again in Section \ref{sec:traversing}).
\textsf{walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$)
both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist, and for allowing access to the memory of the L1 entry via virtual address.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
\looseness=-1
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\ghostmaptoken{\delta{}s}{\mathit{cr3val}}{\delta}) \ast 
  \fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac} \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

While prior work has verified context switches within a single address space~\cite{ni2007contexts}, and context switches
without any code before or after~\cite{syeda2020formal} (i.e., not reasoning about the impact of address space change
on what data was accessible), this is the first verification handling both.

>>>>

\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{\_} \ast \lnot(\textsf{entry\_present entry\_val} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
;;void pte_initialize(pte_t *entry) { ... Stack frame setup elided    
callq  81 <kalloc> ;;allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kalloc}*)
mov    rax,-0x10[rbp] ;; Store into 'local'
;;entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);
mov    -0x10[rbp],rax; mov    -0x8[rbp],rdi; mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \lnot(\textsf{entry\_present entry\_val}) }$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
$\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$
... ;;clean up the stack, return
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\vspace{-1em}
\end{figure}

If the entry is not set, \textsf{pte\_initialize} (Line \ref{line:call_to_pte_initialize} in Figure \ref{fig:calltopteinitialize}) 
allocates a physical page (internally utilizing the only unverified (trusted) code in our case studies, the page-allocator's \textsf{kalloc},\footnote{
  This is an allocator for regions of pre-zeroed physical memory that is mapped, but not accessed by the allocator itself,
  as is typical for slab allocators~\cite{bonwick1994slab}.
  Its verification would be similar to verifying a usermode \textsf{malloc} verifications~\cite{Chlipala2013Bedrock,wickerson2010explicit},
  just with additional invariants on the memory pool.
} 
on Line \ref{line:call_to_kalloc} in Figure \ref{pteinitializespec}). 
Since we are using \textsf{pte\_initialize} for page-table address allocation, we must relate this newly
allocated physical address to the identity mapping map $\Xi$ --- 
see Line \ref{line:page_of_caps} in Figure \ref{fig:calltopteinitialize}, where
\texttt{kalloc}'s specification guarantees it has returned memory from a designated memory
pool that is already mapped
\ifPLDI
\else
\footnote{A reasonable reader might wonder where this pool
initially comes from, and how it might grow when needed. Typically an initial mapping subject to this identity mapping
constraint is set up prior to transition to 64-bit kernel code (notably,
a page table must exist \emph{before} virtual memory is enabled during boot, as part of enabling it is setting
a page table root).
Growing this pool later requires cooperation of physical memory range allocation and virtual memory range allocation,
typically by starting general virtual address allocation at the highest physical memory address plus the identity mapping offset.
This reserves the virtual addresses corresponding to all physical addresses plus the offset for later use in this pool,
as needed.
} 
\fi
and satisfies the offset invariants.
% \todo[inline,color=blue]{colin frontier.
% Stuck with line 31 onwards in Figure 7. rax holds nextpaddr, but I think that should be entrypfn, and 
% the explicit entrypfn id token assertion should go away, as its covered by the forall assertion.
% then the postcondition for pte-initialize should have a specific level now for the entries,
% like 0, which can be updated in the view shift on line 42.
% }
% Focusing on the specification of \textsf{pte\_initialize} separately in Figure \ref{fig:pteinitializespec}, 
% we right immediately realize that instead of seeing see a physical pointsto for the fresly page-table address 
% (e.g. $\mathsf{nextpaddr} \mapsto_{\mathsf{p}} \mathsf{w64\_0}$) deliberately in the post-conditoin in Lines 15-16,
%  we observe a full-ownership token representing the knowledge that a frame and all the entries indexed from this 
% frame are freshly allocated with full-ownership to be a part of the identity map, $\Xi$. 
The soundness argument of this specification relies on the fact that these freshly allocated resources are part 
of an entry construction that has not been completed yet: the presence bit is set 
(Line \ref{line:install_new_entry} in Figure \ref{fig:calltopteinitialize}) after these freshly allocated resources are incorporated to the 
entry construction via the page-frame portion of the PTE. In other words, the side condition, (*),
 formalizes that any access to the entry with these resources is \textit{invalid} (in the sense of not necessarily
having accompanying resources) until the entry is marked present (and thus the memory returned from \textsf{kalloc}
moves into the page table invariant.

@@

\begin{figure}\footnotesize
  \vspace{-1em}
$\begin{array}{l}\left( \begin{array}{l} \ulcorner \textsf{pdp}+\textsf{PDPEX(va)} = \textsf{pdp\_entry} \land \textsf{table\_root(pml4\_entry}_{\textsf{pfn}})  = \textsf{pdp} \\ \land \textsf{pd}+\textsf{PDEX(va)} = \textsf{pd\_entry} \land \textsf{table\_root(pml3\_entry}_{\textsf{pfn}})  = \textsf{pd} \\ \land \textsf{pt}+\textsf{PDPEX(va)} = \textsf{pt\_entry} \land \textsf{table\_root(pml2\_entry}_{\textsf{pfn}})  = \textsf{pt} \urcorner  \ast \\  \textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{pml4\_entry},\textsf{pdp\_entry},\textsf{pd\_entry},\textsf{pte}\},m)  \\ \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast  \\
      \ghostmaptoken{\textsf{id}}{(\mathsf{pml4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{plm2\_entry})}{\textsf{2}}  \ast \\  \textsf{pml4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4e\_val}  \ast \\  \textsf{pdp\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3e\_val}  \ast \\  \textsf{pd\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2e\_val}  \end{array} \right)  =\textsf{R}_{\textsf{walk} } \\ \ast \left(\begin{array}{l} \textsf{rax} \mapsto \textsf{pte} \ast  \textsf{pt\_entry} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present pte\_val})\urcorner \end{array}\right) =\textsf{R}_{\textsf{l1e}}\end{array} $
\caption{\textsf{R}$_{\textsf{walk}}$: Resources obtained from invariant for walking page-table directory with \emph{non-existing entries} via calls to \textsf{pte\_get\_next\_table} in Figure \ref{fig:calltopteinitialize}}
\label{fig:rwalk}
\vspace{-1em}
\end{figure}
Realizing a software page-table walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdir}. 
The special part of the specficication for a page table walk can be considered as accumulation of memory mappings for the page-table entries 
visited and frame addresses for page-tables. 
For example, Lines \ref{line:ex_l4_vpte} and \ref{line:ex_l3_vpte} in Figure \ref{walkpgdir} show the virtual pte-pointsto assertions for L4 and L3 entries.
In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk}}$ -- 
which allows us to construct and return the path to the L1 entry in the tree to insert a new page.

@@

\subsection{Mapping a New Page}
\label{sec:mapnew}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table via a call
\[\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t fpaddr)}\] in Figure \ref{fig:mapping_code}.
To do so, with a given allocated a fresh page (\textsf{fpaddr}), then calculate the appropriate
known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdir} in Figure \ref{fig:mapping_code})  and update 
the appropriate L1 page table entry (Line 35 in Figure \ref{fig:mapping_code});
unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t fpaddr ) {
... ;;setting up the stack      
 $\specline{\textsf{P} \ast \textsf{pml4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$      
;; pte_t *pte = walkpgdir(pml4, a);
 mov  -0x8[rbp], rdi;
 mov  -0x16[rbp], rsi;
 mov  -0x24[rbp], rdx;
;; pte = walkpgdir_succ(pml4, va);
 mov    -0x8[rbp], %rdi;
 mov    -0x16[rbp], rsi;
 mov    0x0, rax;
 callq  291 <walkpgdir> (*\label{line:call_walkpgdir}*)
 $\specline{ \textsf{R}_{\textsf{walk}} \ast \left(\begin{array}{l}   \textsf{rax} \mapsto \textsf{pte} \ast  \textsf{pte} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present pte\_val}) \end{array}\right) =\textsf{R}_{\textsf{l1e}} }$
 mov  rax, rdx;
 mov    rdx,-0x28[rbp];
;;if (!pte->present){ 
 mov    -0x28[rbp], rdx
 mov    [rdx], rcx
 and    0x1, rcx
 mov    rcx, r8
 cmp    0x0, r8
 jne    3e0 <vaspace_mappage+0x190>
 ...
;;pte->pfn          = PTE_ADDR_TO_PFN(pa);
 mov    -0x18[rbp], rdx;
 shr    0xc, rdx;
 mov    -0x28[rbp], rsi;
 mov    [rsi], rdi;
 movabs 0xffffffffff, rcx
 and    rcx, rdx
 shl    0xc, rdx
 movabs 0xfff0000000000fff, rcx
 and    rcx, rdi
 or     rdx, rdi
 mov    rdi, [rsi] (*\label{line:updatepfn}*)
$\specline{ \textsf{pte} \mapsto_{\textsf{vpte,1}} (\textsf{set\_pfn}(\textsf{pte\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present pte\_val}) \urcorner \ast P}$
 ;; Store updated entry back to L1 entry (*\label{line:l1entry_store}*)
;;}
... ;;set present bit in entry, then clean up the stack
... ;;and return either unaccessible pte pointsto or one with fresh page (fpaddr)
 $\specline{ \begin{array}{l}\textsf{pte} \mapsto_{\textsf{vpte,qfrac}} \textsf{pte\_val} \ast \ulcorner (\textsf{entry\_present pte\_val}) \urcorner \ast P \\
 \lor \; \;  \textsf{pte} \mapsto_{\textsf{vpte,1}} (\textsf{set\_pfn}(\textsf{pte\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present pte\_val}) \urcorner \ast P \end{array} }$
;;}
\end{lstlisting}
\vspace{-1em}
  \caption{Specification and proof of code for mapping a new page with $\textsf{R}_{\textsf{walk\_acc}}$ in Figure \ref{walkpgdir} expanded.}
\label{fig:mapping_code}
\end{figure}

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline{walkpgdir} (discussed again in Section \ref{sec:traversing}).
\textsf{walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$)
both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist (R$_{\textsf{walk}}$ in Figure \ref{fig:rwalk}), and for allowing access to the memory of the L1 entry via virtual address (R$_{\textsf{l1e}}$ in Figure \ref{fig:rwalk}).
%A PTE points-to, as we already brifly mentioned, is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
%rather than existentially quantified:
%\looseness=-1
% \[
%\begin{array}{l}
%    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
%    \exists \delta\ldotp
%	(\lambda \mathit{cr3val}\ldotp
%	\ghostmaptoken{\delta{}s}{\mathit{cr3val}}{\delta}) \ast 
%  \fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac} \ast \paddr \mapsto_{\mathsf{p}} \vpage
%\end{array}
%\]
%This supports rules for accessing memory
%at that virtual address, but exposing the physical location being modified
%makes this convenient to use for page table modifications, since we must ensure
%the modified data is at the specific physical location that will be used as the L1 entry.
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

While prior work has verified context switches within a single address space~\cite{ni2007contexts}, and context switches
without any code before or after~\cite{syeda2020formal} (i.e., not reasoning about the impact of address space change
on what data was accessible), this is the first verification handling both.
\looseness=-1
